---
title: "R Coding Exercise"
---

```{r}
# To load the dslabs package
library(dslabs)
# Look at help file for gapminder data
help(gapminder)
# Get an overview of the data structure 
str(gapminder)
# Generate a summary of the data 
summary(gapminder)
```

```{r}
# Determine the type of object gapminder is
class(gapminder)
# Subset the dataset to only include African countries
africadata <- gapminder[gapminder$continent == "Africa", ]
# Check the structure of the new africadata object
str(africadata)
# Check a summary of the new africadata object
summary(africadata)
```

```{r}
# Determine the type of object gapminder is
class(gapminder)
# Subset the dataset to only include African countries
africadata <- gapminder[gapminder$continent == "Africa", ]
# Check the structure of the new africadata object
str(africadata)
# Check a summary of the new africadata object
summary(africadata)
```

```{r}
# Create an object containing only the 'infant_mortality' variable
africadata_infant <- africadata[, "infant_mortality"]
# Check the structure of the new object
str(africadata_infant)
# Check a summary of the new infant mortality object
summary(africadata_infant)
```

```{r}
# Create an object containing only the 'infant_mortality' variable
africadata_infant <- africadata[, "infant_mortality"]
# Check the structure of the new object
str(africadata_infant)
# Check a summary of the new infant mortality object
summary(africadata_infant)
```

```{r}
# Create an object containing only the 'life_expectancy' variable
africadata_life <- africadata[, "life_expectancy"]
# Check the structure of the new life ex. object
str(africadata_life)
# Check a summary of the new life ex. object
summary(africadata_life)
```

```{r}
# Create an object containing only the 'life_expectancy' variable
africadata_life <- africadata[, "life_expectancy"]
# Check the structure of the new life ex. object
str(africadata_life)
# Check a summary of the new life ex. object
summary(africadata_life)
```

```{r}
# Load the plotting tools library through ggplot package
library(ggplot2)

# Plot life expectancy as a function of infant mortality + theme tool is something found in use of AI for my plots 
ggplot(africadata, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +  

# Plot above data as points
  labs(
    title = "Life Expectancy vs. Infant Mortality",
    x = "Infant Mortality (per 1000 live births)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()
  
# Plot life expectancy as a function of population size with log scale on x-axis, using the + theme tool is something found in use of AI for my plots 
ggplot(africadata, aes(x = population, y = life_expectancy)) +
  geom_point() +  
# Plot above data as points
  scale_x_log10() +  
# Set x-axis to log scale
  labs(
    title = "Life Expectancy vs. Population Size",
    x = "Population Size (log scale)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()
```

## The data streaks are likely due to the large number of data plots that are close in relative number, especially on a log scale. This makes the data appear to be in streaks rather than in individual points. 

```{r}
# Identify rows where infant_mortality is missing (NA)
missing_infant_mortality <- africadata[is.na(africadata$infant_mortality), ]
# Extract the unique years with missing infant mortality data
missing_years <- unique(missing_infant_mortality$year)
# Print the result
missing_years
# Subset the africadata object to only include data for the year 2000
africadata_2000 <- africadata[africadata$year == 2000, ]
# Check the structure of the new object
str(africadata_2000)
# Check a summary of the new object
summary(africadata_2000)
```

```{r}
# Load necessary libraries again 
library(ggplot2)

# Make Plot 1: Life Expectancy vs. Infant Mortality for the year 2000
ggplot(africadata_2000, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +  

# Plot data as points
  labs(
    title = "Life Expectancy vs. Infant Mortality (Year 2000)",
    x = "Infant Mortality (per 1000 live births)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()

# Plot 2: Life Expectancy vs. Population Size (Log scale) for the year 2000
ggplot(africadata_2000, aes(x = population, y = life_expectancy)) +
  geom_point() +  

# Plot data as points
  scale_x_log10() +  

# Set x-axis to log scale
  labs(
    title = "Life Expectancy vs. Population Size (Year 2000)",
    x = "Population Size (log scale)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()
```


```{r}
# Fit a simple linear model with life expectancy as the outcome and infant mortality as the predictor
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata_2000)
# Apply the summary to fit1 to display the results
summary(fit1)
# Fit a linear model with life expectancy as the outcome and population size as the predictor
fit2 <- lm(life_expectancy ~ population, data = africadata_2000)
# Apply the summary to fit2 to display the results
summary(fit2)
```

# Based on the data's generated significantly small p scores, there is highly statistically significant relationships between both life expectancy and infant mortality as well as life expectancy and population size. This makes intuitive sense based on what is known about demographic transition and development, which is under the purview of the gapminder data on developing African countries. 

# Alexis Gonzalez contributed to the following section!

### looking at help file for greenhouse_gases data

```{r}
library(dslabs)
help(greenhouse_gases)
```

### checking out the data set

```{r}
str(greenhouse_gases)
summary(greenhouse_gases)
```

### Checking what type of object greenhouse_gases is

```{r}
class(greenhouse_gases)
```

### There aren't many variable in this but I want to see just years

```{r}
library(tidyverse)
grnhsyr <- greenhouse_gases %>%
  select(year)
str(grnhsyr)
summary(grnhsyr)
```

```{r}
library(ggplot2)
```

### now lets make a variable that summarizes the concentration of all gases each year so we can see gas concetration over years. Later we can look at each gas

```{r}
#first we have to summarize total concentration by year
total_by_year <- greenhouse_gases %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(total_concentration = sum(concentration))
```

### Now lets look for some trends in gas concentration over years

```{r}
p1 <- ggplot(total_by_year, aes(x=year, y=total_concentration)) +
geom_line(size = 1, color = "lightblue")
  title = "Gas concentration over Years"
  x= "years"
  y= " Total concentration"
print(p1)
```

### We see that the concentration of greenhouse gases skyrocket in 20-21st century, lets see the concentration of gases from just that time period

```{r}
#make a new variable for desired time frame
gasyears <- greenhouse_gases %>%
  filter(year >= 1900 & year <= 2000) %>%
  group_by(year) %>%
  summarize(total_concentration = sum(concentration))
  
p2 <- ggplot(gasyears, aes(x=year, y=total_concentration)) +
geom_line(size = 1, color = "darkgreen")
  title = "Gas concentration from 1900 to 2000"
  x= "years"
  y= " Total concentration"
print(p2)
```

### To no surprise we see the highest amount of greenhouses gases in the year 2000 so finally lets see a break down of the gas concentration by gas

```{r}
#first I'm going to filter the data to only include the year 2000
gas_2000 <- greenhouse_gases %>%
  filter(year == 2000)
```

```{r}
# Now I'm going to make a bar graph
p3 <- ggplot(gas_2000, aes( x = gas, y = concentration, fill = gas))+
  geom_bar(stat = "Identity") + 
  labs(
    title = "Greenhouse Gases in 2000",
    x = "Greenhouse gas",
    y = "Concentration"
    
  )
print(p3)
```

### Finally we will create a linear model to see the relationship between time in years and gas concentration

```{r}
model <- lm(total_concentration ~ year, data=total_by_year)
summary(model)
p4 <- ggplot(total_by_year, aes(x= year, y= total_concentration)) +
  geom_point(size = 3, color = "aquamarine3" ) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "relationship between year and gas concentration"
  )
print(p4)
```

### From this analysis we see that year and gas concentration have a positive relationship, as the year increases total gas concentration increases. There is an exponential increase in gas concentration around the year 1900, peaking at 2000. In 2000 the most abundant greenhouse gas was CH4 or methane.

